```{r 07_setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE, fig.align="center")
```

# IP weighting for censoring

## Learning Goals {-}

1. Understand how IP weighting can be used to address the lack of generalizability caused by censoring

<br><br><br><br>


## Discussion {-}

**Language for interpreting coefficients in MSMs**

$$ E[Y^a] = \beta_0 + \beta_1 a $$

- Context: Y is cholesterol, A is medication (a = 1: on medication)
- $\beta_0$: The average potential cholesterol level when all are not on medication
- $\beta_1$: The average causal effect: the change in average potential cholesterol level if everyone were on medication compared to if no one were on medication


<br><br><br><br>


**Do we really care about the ACE?**

- Depends on research goals, but yes, sometimes ACE is of interest.
    - Is it a nationwide policy? If so, then enacting the policy means "treating" everyone and not enacting means treating no one.
- May care about the ACE within subgroups. We've seen this with effect modfication MSMs:

$$ E[Y^a] = \beta_0 + \beta_1 a + \beta_2 M + \beta_3 a\times M $$

- Say $M$ indicates history of heart disease (1: history. 0: no history)
- **Question:** What are the interpretation of $\beta_1$ and $\beta_3$ in this model?


<br><br><br><br>


**What is the difference between regular and IP-weighted GLMs?**

- Our tree diagrams showed that weighting allows us to create "populations" of all treated and all untreated.
    - Our WHATIF book calls the combined population of the all treated and the all untreated a **pseudopopulation**
- We can model the outcomes in this pseudopopulation with standard regression models (marginal structural models).
    - Allows us to estimate causal quantities: $P(Y \mid \hbox{do}(A=a))$, $E[Y \mid \hbox{do}(A=a)]$ (in potential outcome notation: $P(Y^a)$, $E[Y^a]$)
    - Standard regression models without IP weights do not estimate these causal quantities.


<br><br><br><br>


**Censoring and selection bias**

Reminder: NHEFS investigation from last time

> **Research goal:** What is the average causal effect of smoking cessation on weight gain at a follow-up visit about 10 years later?

- There were some people who did not have their weight measured at Visit 2. They must be excluded from our analysis.
    - Define a censoring indicator $C$ where $C = 1$ means censored/excluded and $C = 0$ means uncensored/included.
    - Our analysis only generalizes to uncensored individuals--the types of people who would stay in the study!

```{r echo=FALSE, eval=TRUE, fig.width=15, fig.height=6}
library(dagitty)
dag <- dagitty("dag {
bb=\"0,0,1,1\"
C [adjusted,pos=\"0.453,0.521\"]
Personality [latent,pos=\"0.316,0.742\"]
VidGames [pos=\"0.639,0.589\"]
active [pos=\"0.632,0.207\"]
age [pos=\"0.214,0.207\"]
education [pos=\"0.338,0.207\"]
exercise [pos=\"0.700,0.207\"]
qsmk [exposure,pos=\"0.139,0.484\"]
race [pos=\"0.278,0.207\"]
sex [pos=\"0.146,0.207\"]
smokeintensity [pos=\"0.442,0.206\"]
smokeyrs [pos=\"0.560,0.206\"]
wt71 [pos=\"0.762,0.206\"]
wt82_71 [outcome,pos=\"0.828,0.484\"]
Personality -> C
Personality -> VidGames
VidGames -> wt82_71
active -> exercise
active -> qsmk
active -> wt82_71
age -> qsmk
age -> wt82_71
education -> qsmk
education -> wt82_71
exercise -> qsmk
exercise -> wt82_71
qsmk -> C
qsmk -> wt82_71
race -> qsmk
race -> wt82_71
sex -> qsmk
sex -> wt82_71
smokeintensity -> qsmk
smokeintensity -> smokeyrs
smokeintensity -> wt82_71
smokeyrs -> qsmk
smokeyrs -> wt82_71
wt71 -> qsmk
wt71 -> wt82_71
}
")
plot(dag)
```

- Before: just wanted to find $Z$ such that conditional on $Z$ (within subsets defined by $Z$), the outcomes of the treated and untreated were comparable.
    - Could appropriately upweight the treated and untreated to get "populations" of all treated and all untreated.

<br>

- Now: also want to ensure that the outcomes in the censored and uncensored are comparable.
    - Can appropriately upweight the uncensored to represent everyone had they not been censored.
    - The censored get zero weight. (We don't observe them!)

<br>

- When are the outcomes in the censored and uncensored comparable?
    - When variables in $A$ and $Z_2$ together d-separate $C$ and the outcome $Y$.
    - e.g., The DAG below. View $C$ as the new "treatment" variable.
    - Another way to put it: when we put a box around $C$ in the DAG, $Z$ must d-separate $A$ and $Y$ under the null.    
    Same condition as before, now explicit about paths with conditioned-on colliders (non-backdoor paths).

```{r echo=FALSE, eval=TRUE, fig.width=15, fig.height=6}
plot(dag)
```

**Questions:** $C$ is inherently conditioned on in our analysis ($C = 0$).

- What set $Z$ d-separates the treatment and outcome under the null?
- What is a set $Z_2$ such that $Z_2 \cup A$ (read $Z_2$ "union" $A$, or $Z_2$ and $A$ together) d-separate $C$ and the outcome?


<br><br><br><br>


**Conclusion:**

- $Z$ should include:    
    `sex`, `age`, `race`, `education`, `smokeintensity`, `smokeyrs`, `active`, `exercise`, `wt71`, and `VidGames`    
    (Same as before with the addition of `VidGames`)
- $Z$ blocks spurious, non-causal paths between $A$ (`qsmk`) and $Y$ (`wt82_71`)

<br>

- $Z_2$ should include `VidGames`
- Together with $A$, $Z_2$ blocks spurious, non-causal paths between $C$ and $Y$
- Note that $Z_2$ is contained within $Z$.
    - This will be the case when $A$ is a cause of $C$ (often true).


<br><br><br><br>


**Ultimate goal:** Want to upweight individuals using $P(A, C = 0 \mid Z)$

- Why? Same rationale as our tree diagram with additional $C = 1$ and $C = 0$ branches.
- Allows us to estimate $E[Y^{a, c = 0}]$: expected potential outcome under treatment $a$ and when no one has been censored ($C = 0$)


<br><br><br><br>


We can express $P(A, C = 0 \mid Z)$ as the following:

$$
\begin{align*}
P(A, C = 0 \mid Z) &= P(A, C = 0, Z)/P(Z) \\
&= P(A \mid Z) P(C = 0 \mid A, Z)
\end{align*}
$$

- Based on the last line, we see (1) the propensity score and (2) another "propensity" of being uncensored.
- Will want models for both.
- (The censoring probability is not typically called a propensity score.)


<br><br><br><br>


A little extra theory (don't worry about the math part below if you haven't taken Probability):

$Z$ may contain extra variables not in $Z_2$ if there are, say, many common causes of $A$ and $Y$, but only a few of these are a cause of $C$. Let's say that these extra variables are in the set $Z_1$. ($Z = Z_1 \cup Z_2$)

If the variables in $Z_1$ are not causes of $C$, then $C$ and $Z_1$ are conditionally independent given $A$ and $Z_2$.

Going through the work below...

$$
\begin{align*}
P(A, C = 0 \mid Z) &= P(A, C = 0 \mid Z_1, Z_2) \\
&= P(A, C = 0, Z_1, Z_2)/P(Z_1, Z_2) \\
&= P(A \mid Z_1, Z_2) P(C = 0 \mid A, Z_1, Z_2) \\
&= P(A \mid Z_1, Z_2) P(C = 0 \mid A, Z_2)
\end{align*}
$$

...this implies that the propensity score model (for $A$) should depend on all variables in $Z$ (the set $Z_1$ and $Z_2$ together) and that the censoring probability model only need depend on $Z_2$.


<br><br><br><br>


**Analysis plan:**

Notation:

- $Z$ is the set that d-separates $A$ and $Y$ when $C$ is conditioned on.
- $Z_2$ (a subset of $Z$) that, together with $A$, d-separates $C$ and $Y$

1. Model the propensity scores $P(A\mid Z)$ using all variables $Z$ in the d-separating set. Create inverse probability weights $W^A = 1/P(A\mid Z)$ (treatment weights).

2. Model the probability of censoring $P(C \mid A, Z_2)$. Create inverse probability weights $W^C = 1/P(C = 0\mid A, Z_2)$ (censoring weights).

3. Create final weights (for treatment and censoring) $W^{A,C} = W^A\times W^C$.

4. Fit the desired marginal structural model with these final weights.

<br><br><br><br>


## Exercises {-}

A template Rmd is available [here](template_rmds/07-ipw-censoring.Rmd).

### Setup {-}

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(geepack)

nhefs <- read_csv("https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/1268/20/nhefs.csv")
```

Create a censoring indicator variable that will be `TRUE` if the individual was censored and `FALSE` otherwise.

```{r}
nhefs$cens <- is.na(nhefs$wt82_71)
```

Throughout we'll use the full `nhefs` dataset, and we'll work from the DAG below ($C$ has been conditioned on):

```{r echo=FALSE, eval=TRUE, fig.width=15, fig.height=6}
dag2 <- dagitty("dag {
bb=\"0,0,1,1\"
C [adjusted,pos=\"0.453,0.521\"]
Personality [latent,pos=\"0.268,0.107\"]
active [pos=\"0.632,0.207\"]
age [pos=\"0.214,0.207\"]
education [pos=\"0.338,0.207\"]
exercise [pos=\"0.700,0.207\"]
qsmk [exposure,pos=\"0.139,0.484\"]
race [pos=\"0.278,0.207\"]
sex [pos=\"0.146,0.207\"]
smokeintensity [pos=\"0.442,0.206\"]
smokeyrs [pos=\"0.560,0.206\"]
wt71 [pos=\"0.762,0.206\"]
wt82_71 [outcome,pos=\"0.828,0.484\"]
Personality -> education
active -> exercise
active -> qsmk
active -> wt82_71
age -> C
age -> qsmk
age -> wt82_71
education -> C
education -> qsmk
education -> wt82_71
exercise -> qsmk
exercise -> wt82_71
qsmk -> C
qsmk -> wt82_71
race -> qsmk
race -> wt82_71
sex -> qsmk
sex -> wt82_71
smokeintensity -> qsmk
smokeintensity -> smokeyrs
smokeintensity -> wt82_71
smokeyrs -> qsmk
smokeyrs -> wt82_71
wt71 -> qsmk
wt71 -> wt82_71
}")
plot(dag2)
```

<br><br>

### Part 1: Modeling to obtain weights {-}

a. Based on the DAG above, identify the set of variables $Z$ that d-separates `qsmk` and weight gain.

b. Fit an appropriate treatment propensity score model, and call this `ps_treat_mod`. For simplicity, assume that a quadratic relationship for the quantitative variables gives a good fit. Make sure to wrap categorical variables inside `factor()` in your model formula.

c. Add a new variable to your dataset called `weights_treat` that contains the IP weights for treatment.

d. Is it believable that quitting smoking (treatment) is a cause of censoring? How do the tabulations/calculations below help answer this?    
    ```{r}
    # P(censored | quitters) and P(censored | nonquitter)
    table(qsmk = nhefs$qsmk, cens = nhefs$cens)
    table(qsmk = nhefs$qsmk, cens = nhefs$cens) %>% prop.table(margin = 1)
    ```

e. How would you see if the data support the `age --> C` and `education --> C` arrows? Describe, but don't actually perform this analysis.

f. Fit an appropriate model for the probability of censoring, and call this `prob_cens_mod`. For simplicity, assume that a quadratic relationship for the quantitative variables gives a good fit. Make sure to wrap categorical variables inside `factor()` in your model formula.

g. Add a new variable to your dataset called `weights_cens` that contains the IP weights for censoring. (Note that in R, the default for logistic regression is to model the probability of the outcome variable equaling 1.)

h. Add a final weight variable to your dataset called `weight_TC`.

<br><br><br>

### Part 2: Fitting MSMs {-}

Fit the following two MSMs ($A$ is `qsmk`, and for `sex`, 0 indicates males and 1 indicates females):

$$ E[Y^{a, c=0}] = \beta_0 + \beta_1 a $$
$$ E[Y^{a, c=0}] = \beta_0 + \beta_1 a + \beta_2\hbox{sex} + \beta_3 a\times\hbox{sex} $$

Note: In our previous analysis where we worked with `nhefs_subs`, we had actually fit the following:

$$ E[Y^{a} \mid C = 0] = \beta_0 + \beta_1 a $$
$$ E[Y^{a} \mid C = 0] = \beta_0 + \beta_1 a + \beta_2\hbox{sex} + \beta_3 a\times\hbox{sex} $$

Refit these models here using the relevant weights. How do your results compare?






