```{r 09_setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE)
```

# (PART) Sensitivity Analyses {-}

# Sensitivity Analyses for Unmeasured Confounding

## Learning Goals {-}

1. Understand the role of sensitivity analyses in sound scientific practice
2. Implement and interpret the results of sensitivity analyses

<br><br>

## Exercises {-}

A template Rmd is available [here](template_rmds/09-sensitivity.Rmd).

```{r}
library(ggplot2)
library(dplyr)
library(geepack)
```

Given the code below, what is the underlying DAG?

```{r}
set.seed(302)
n <- 5000
Z <- rbinom(n, size = 1, prob = 0.5)
p_A <- dplyr::case_when(
    Z==1 ~ 0.6,
    Z==0 ~ 0.2
)
A <- rbinom(n, size = 1, prob = p_A)
Y <- 0.1*A + 0.3*Z + rnorm(n)

sim_data <- data.frame(A, Y, Z)

# Create a subject ID variable
sim_data$subj_id <- seq_len(n)
```

First *outline* the process for fitting a marginal structural model to estimate the average causal effect of treatment $A$ on outcome $Y$. Then write the relevant code. You will need to use `subj_id` as the `id` variable within `geeglm()`.

```{r}
# Original MSM fit

```

Here we begin our sensitivity analysis. We'll create a binary unmeasured variable $U$ that has varying degrees of association with both $A$ and $Y$.

- Add extra comments throughout the code to better document this implementation of a sensitivity analysis.
- Complete the sensitivity analysis by completing the `TASK` referred to within the for-loop.

```{r}
# Create a vector of possible associations that U has with A and Y
U_assocs <- seq(0.05,2,0.1)

# Create empty storage containers to store the new ACEs and CI bounds
new_aces <- rep(0, length(U_assocs))
new_ci_upper <- rep(0, length(U_assocs))
new_ci_lower <- rep(0, length(U_assocs))

# Loop over possible values of the association
for (i in seq_along(U_assocs)) {
    U_assoc <- U_assocs[i]
    log_odds_U <- U_assoc*sim_data$Y + U_assoc*sim_data$A
    p_U <- exp(log_odds_U)/(1+exp(log_odds_U))
    U <- rbinom(n, size = 1, prob = p_U)
    sim_data$U <- U

    # TASK: Fit the new MSM that accounts for U
    new_msm_fit <- ???

    # Store relevant values
    new_aces[i] <- summary(new_msm_fit)$coefficients["A", "Estimate"]
    se <- summary(new_msm_fit)$coefficients["A", "Std.err"]
    new_ci_upper[i] <- new_aces[i]+(qnorm(0.975)*se)
    new_ci_lower[i] <- new_aces[i]-(qnorm(0.975)*se)
}
```

Below we plot the results of this sensitivity analysis. The black line shows the different ACEs, and the ribbon shows the corresponding 95% confidence interval. The red line shows the original ACE estimate.

- What information can be gained from this plot?
- What is the smallest magnitude of association with $U$ that leads to qualitatively different conclusions?
- Look back to where we simulated $Y$. How could we change this so that the smallest magnitude referred to in the previous question is bigger?

```{r}
sens_data <- data.frame(lnOR = U_assocs, ace = new_aces, ci_lower = new_ci_lower, ci_upper = new_ci_upper)
ggplot(sens_data, aes(x = lnOR, y = ace)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), fill = "grey70") +
    geom_line() +
    geom_hline(aes(yintercept = summary(orig_msm_fit)$coefficients["A", "Estimate"]), color = "red")
```

**Closing questions:** How could we adapt this structure for a sensitivity analysis to situations where the variables under consideration are of different types (quantitative vs. categorical) than what we considered here? How could these sensitivity analyses be used for assessing the impact of even more general structures of unmeasured confounding?
