---
title: "Introductions, Review, and Motivation"
author: "Your Name"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Setup

```{r}
# Read in the data
chemo_study_data <- read.csv("https://www.dropbox.com/s/vl06j75a8afw8ct/chemo_study.csv?dl=1")

# Load the ggplot2 package
```

# Exercise 1

The first step in any data analysis is to visualize your data. Let's refamiliarize ourselves with the `ggplot2` package in R. It may be helpful to have [this ggplot2 cheat sheet](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) open.

## Part a

Look at the distribution of each of the 4 measured variables. What plot type is most appropriate for this type of variable?

```{r}
ggplot(chemo_study_data, aes(x = pre_tumor_size)) + ???
```

## Part b

Is pre-treatment tumor size predictive of whether or not a patient received chemotherapy? Make a plot to assess this, and briefly state what conclusions can be drawn from the plot. (Hint: it will be helpful to look at the second page of the cheat sheet in the section labeled "Position Adjustments".)

```{r}
ggplot(chemo_study_data, aes(x = pre_tumor_size, fill = treated)) + ???
```

## Part c

Is pre-treatment tumor size predictive of whether or not a patient recovered? Make a plot, and briefly state your conclusions.

## Part d

A variable is a **confounder** if it is a common cause of both the treatment and outcome. This is shown in the diagram below. Given your results from parts b and c, could pre-treatment tumor size be a confounder of the relationship between chemotherapy treatment and recovery? If yes, what is the concern here?





Note: the causal relationships of interest so far can be depicted in a **causal diagram**, shown below. An arrow between two variables indicates that one is a cause of the other (an arrow points from a cause to its effect).

```{r fig.width=6, fig.height=2, fig.align="center"}
# If you're curious about how to draw the causal diagram in R
par(mar = rep(0,4))
plot(1, type = "n", xaxt = "n", yaxt = "n", bty = "n", xlab = "", ylab = "", xlim = c(0,6), ylim = c(1,4))
text(c("Chemotherapy", "Pre-treatment tumor size", "Recovery"), x = c(1,3,5), y = c(1,4,1), cex = 1.1)
arrows(x0 = c(3,3), y0 = c(4,4)-0.2, x1 = c(1,5), y1 = c(1,1)+0.2, angle = 25, lwd = 4)
arrows(x0 = 1.7, y0 = 1, x1 = 4.5, y1 = 1, angle = 25, lwd = 4)
```





# Exercise 2

One way to "adjust for" the influence of counfounders is to include them as predictors/explanatory variables in regression models.

We can model recovery using a **logistic regression model** (used when the outcome variable is binary). In R, we can fit a logistic regression model using code like the following:

```{r eval=FALSE}
# Fit the model and store it in the "mod" object
mod <- glm(outcome_variable ~ explanatory_variable1+explanatory_variable2,
    family = "binomial", data = your_data)
# Display model output
summary(mod)
```

## Part a

Fit a logistic regression model with only treatment as a predictor. Interpret the treatment coefficient. Is this the interpretation you expected?

## Part b

Fit a logistic regression model with both treatment and pre-treatment tumor size as predictors. Interpret the treatment coefficient. Is this the interpretation you expected?





# Exercise 3

So far, we have not considered the `post_tumor_size` variable at all in our analyses.

## Part a

Let's get some visual understanding of this variable. Make 2 plots: one showing its relationship with the `treated` variable and the second showing its relationship with the `recovery` variable.

## Part b

What is your intuition - should we include `post_tumor_size` as an explanatory variable in our logistic regression model? Why or why not?

## Part c

Fit the logistic regression model (`mod3`) with all 3 explanatory variables, and interpret the treatment coefficient. How do the results compare to `mod2`?

## Part d

The diagram that we considered at the end of Exercise 1 only included the treatment, recovery, and pre-treatment tumor size variables. Draw (on paper) an expanded causal diagram that includes post-treatment tumor size. In light of where post-treatment tumor size appears on your diagram, what does it mean to include it in the logistic regression model?

## Part e

The data in this investigation were simulated. That is, your instructor generated them, and thus, she actually knows the truth behind the data. The truth is that chemotherapy does have a true beneficial causal impact on recovery. Chemotherapy does increases the chance of recovery. Given this information and the insights you gained from these three exercises, try to put all of this information together. What makes sense? What remains unclear?






