---
title: "IPW in practice"
output: 
  html_document:
    toc: true
    toc_float: true
---

## Setup

The dataset codebook is available [here](https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/2012/10/NHEFS_Codebook.xls).

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(geepack)

nhefs <- read_csv("https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/1268/20/nhefs.csv")
```

Create a censoring indicator variable that will be `TRUE` if the individual was censored and `FALSE` otherwise.

```{r}
nhefs$cens <- is.na(nhefs$wt82_71)
```

Throughout we'll use the full `nhefs` dataset, and we'll work from the DAG below ($C$ has been conditioned on):

```{r fig.width=15, fig.height=6}
# Install the dagitty package if you want this to display
library(dagitty)
dag2 <- dagitty("dag {
bb=\"0,0,1,1\"
C [adjusted,pos=\"0.453,0.521\"]
Personality [latent,pos=\"0.268,0.107\"]
active [pos=\"0.632,0.207\"]
age [pos=\"0.214,0.207\"]
education [pos=\"0.338,0.207\"]
exercise [pos=\"0.700,0.207\"]
qsmk [exposure,pos=\"0.139,0.484\"]
race [pos=\"0.278,0.207\"]
sex [pos=\"0.146,0.207\"]
smokeintensity [pos=\"0.442,0.206\"]
smokeyrs [pos=\"0.560,0.206\"]
wt71 [pos=\"0.762,0.206\"]
wt82_71 [outcome,pos=\"0.828,0.484\"]
Personality -> education
active -> exercise
active -> qsmk
active -> wt82_71
age -> C
age -> qsmk
age -> wt82_71
education -> C
education -> qsmk
education -> wt82_71
exercise -> qsmk
exercise -> wt82_71
qsmk -> C
qsmk -> wt82_71
race -> qsmk
race -> wt82_71
sex -> qsmk
sex -> wt82_71
smokeintensity -> qsmk
smokeintensity -> smokeyrs
smokeintensity -> wt82_71
smokeyrs -> qsmk
smokeyrs -> wt82_71
wt71 -> qsmk
wt71 -> wt82_71
}")
plot(dag2)
```






## Part 1: Modeling to obtain weights

a. Based on the DAG above, identify the set of variables $Z$ that d-separates `qsmk` and weight gain.

b. Fit an appropriate treatment propensity score model, and call this `ps_treat_mod`. For simplicity, assume that a quadratic relationship for the quantitative variables gives a good fit. Make sure to wrap categorical variables inside `factor()` in your model formula.

c. Add a new variable to your dataset called `weights_treat` that contains the IP weights for treatment.

d. Is it believable that quitting smoking (treatment) is a cause of censoring? How do the tabulations/calculations below help answer this?    

```{r}
# P(censored | quitters) and P(censored | nonquitter)
table(qsmk = nhefs$qsmk, cens = nhefs$cens)
table(qsmk = nhefs$qsmk, cens = nhefs$cens) %>% prop.table(margin = 1)
```

e. How would you see if the data support the `age --> C` and `education --> C` arrows? Describe, but don't actually perform this analysis.

f. Fit an appropriate model for the probability of censoring, and call this `prob_cens_mod`. For simplicity, assume that a quadratic relationship for the quantitative variables gives a good fit. Make sure to wrap categorical variables inside `factor()` in your model formula.

g. Add a new variable to your dataset called `weights_cens` that contains the IP weights for censoring. (Note that in R, the default for logistic regression is to model the probability of the outcome variable equaling 1.)

h. Add a final weight variable to your dataset called `weight_TC`.





## Part 2: Fitting MSMs

Fit the following two MSMs ($A$ is `qsmk`, and for `sex`, 0 indicates males and 1 indicates females):

$$ E[Y^{a, c=0}] = \beta_0 + \beta_1 a $$
$$ E[Y^{a, c=0}] = \beta_0 + \beta_1 a + \beta_2\hbox{sex} + \beta_3 a\times\hbox{sex} $$

Note: In our previous analysis where we worked with `nhefs_subs`, we had actually fit the following:

$$ E[Y^{a} \mid C = 0] = \beta_0 + \beta_1 a $$
$$ E[Y^{a} \mid C = 0] = \beta_0 + \beta_1 a + \beta_2\hbox{sex} + \beta_3 a\times\hbox{sex} $$

Refit these models here using the relevant weights. How do your results compare?






