---
title: "Probability and DAGs"
author: "Your Name"
output: 
  html_document:
    toc: true
    toc_float: true
---

# Loading packages

```{r}
library(ggplot2)
library(dplyr)
```

# Discussion

An introduction to the `rbinom` function:

```{r}
# 4 different people each flip a fair coin once
rbinom(4, size = 1, prob = 0.5)

# 4 different people flip loaded coins
# First 2 flip a coin with P(Heads) = 0.9
# Second 2 flip a coin with P(Heads) = 0.2
rbinom(4, size = 1, prob = c(0.9, 0.9, 0.2, 0.2))
```

```{r}
# Before you run this code, what will the result be?
rbinom(2, size = 1, prob = c(1, 0))
```

How can we simulate X -> Y for binary X and Y?

```{r}
# set.seed() ensures reproducible random numbers
set.seed(2020)
# Set sample size
n <- 1e5
# Generate a binary variable X for all n cases
X <- rbinom(n, size = 1, prob = 0.5)
Y <- rbinom(n, size = 1, prob = 0.5)
```

Are the X and Y we just simulated independent or dependent? Let's make a plot to see:

```{r}
# We put X and Y in a dataset using data.frame()
# They are turned into categorical variables using factor()
sim_data <- data.frame(X = factor(X), Y = factor(Y))

# Proportional bar plot
ggplot(sim_data, aes(x = X, fill = Y)) +
    geom_bar(position = "fill")
```

How can we make Y dependent on X?

```{r}
# Generate probabilities that Y=1 that depend on the value of X
p_Y <- dplyr::case_when(
    X==1 ~ 0.8,
    X==0 ~ 0.3
)

# Simulate Y using p_Y
Y <- rbinom(n, size = 1, prob = p_Y)

# Put X, Y, and p_Y into a dataset
sim_data <- data.frame(X = factor(X), Y = factor(Y), p_Y = p_Y)

# Check that p_Y = 0.8 when X is 1
sim_data %>%
    filter(X=="1")

# Check that p_Y = 0.3 when X is 0
sim_data %>%
    filter(X=="0")

# Proportional bar plot
ggplot(sim_data, aes(x = X, fill = Y)) +
    geom_bar(position = "fill")
```

How can we make a variable Z dependent on 2 causes: X and Y?

```{r}
X <- rbinom(n, size = 1, prob = 0.5)
Y <- rbinom(n, size = 1, prob = 0.5)
p_Z <- dplyr::case_when(
    X==1 & Y==1 ~ 0.8,
    X==0 & Y==1 ~ 0.3,
    X==1 & Y==0 ~ 0.5,
    X==0 & Y==0 ~ 0.9
)
Z <- rbinom(n, size = 1, prob = p_Z)

# Put X, Y, and Z into a dataset
sim_data <- data.frame(X = factor(X), Y = factor(Y), Z = factor(Z))

# If we wanted to save datasets filtered on Z (Why?)
sim_data_filt_z0 <- sim_data %>% filter(Z=="0")
sim_data_filt_z1 <- sim_data %>% filter(Z=="1")
```





# R Exercises

## Exercise 1

Simulate data for a chain X -> Y -> Z where all variables are binary. Make plots to show the following properties:

- X and Z are marginally dependent
- X and Z are conditionally independent given Y

Describe how your simulation would change if you wanted to check the more general property of chains that two variables are conditionally independent given any variables in between them. (You don't actually implement this simulation but can if you have time.)



## Exercise 2

Simulate data for a fork Y <- X -> Z where all variables are binary. Make plots to show the following properties:

- Y and Z are marginally dependent
- Y and Z are conditionally independent given X



## Exercise 3

Simulate data for a collider X -> Z <- Y where all variables are binary. Make plots to show the following properties:

- X and Y are marginally independent
- X and Y are conditionally dependent given Z

Describe how your simulation would change if you wanted to check the more general property of colliders that X and Y become dependent conditional on Z and/or any descendants of Z. (You don't actually implement this simulation but can if you have time.)



# Conceptual Exercises

(See course website for images in this part.)

## Exercise 4

In the causal diagram above, let X be a treament and Y be an outcome of interest.

a. What would you say are the "causal paths" between X and Y? What would you say are the "non-causal paths"?
b. Thinking about the marginal and conditional independence and dependence relations we've discussed, how might we restrict the influence of these non-causal paths?



## Exercise 5

Can you come up with a general rule that tells us when two variables will be conditionally independent given another set of variables Z? (Where it is possible that Z is an empty set containing no variables.)

Try out your rule on the causal diagram below.

